Sampling requirements to detect ecosystem change

### 2020-01-15 R/manuscript_draft.Rmd
add conceptual figure to draft and edit initial script to make it work (source function code mainly)
updates to introduction and formatting of article

---
title: "Sampling requirements to detect ecosystem change"
author: "Rosalie Bruel and Easton White"
date: "03/10/2019"
output:
  html_document:
    df_print: paged
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
fontsize: 11pt
editor_options:
  chunk_output_type: console
  df_print: paged
---

@@

@@

- writing intro
- writing methods/results pair of section
- writing dicussion
- setting up bibliography system
- Possible journals: Methods Ecology and Evolution, Biological Conservation, *Journal of Environmental Management*, *Environmental Monitoring and Assessment*, BioScience

@@

- Set up ideas about monitoring and cost of monitoring

Environmental monitoring is one of the core components to modern research and management (citations). However, many monitoring programs are expensive and time-consuming (citations). Therefore, an optimal...

- Idea of detecting change over time

One of the primary goals of monitoring programs is to detect long-term ecological change...

- Historical data and ability to choose how much to sample

Because of new technological advances, there are many data sources that can be derived long after the actual  processes occured. For example, paleoecological lake cores can be extracted and ecological communities can be built from this... Similarly, samples for eDNA or videographical approaches can record time points of a system and be analyzed later (citations). In each of these cases, decisions have to made about how much data to extract from the previously collected samples. Should the paleo core be analyzed at every centimeter? Should the video be assessed at every minute? As long as analyzing samples is expensive, these tradeoffs will remain.

Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where additional samples can be added. We first investigate these tools using a simulated-based apporach. We then apply these tools to the detection of change points for a set of lakes...

@@

<img src="Fig_conceptual.pdf" alt="conceptual figure"  width="4200" height="6200">

@@

We began with a theoretical exploration of the determiners of sampling required to detect a changepoint. We modeled a simple AR-1 process with a response variable that represents either population size, biodiversity, or some metric of community composition. The model includes temporal variability ($\sigma$), autocorrelation ($\phi$), the size of the shift ($\delta$), and a specific time when the shift occured ($\tau$):

@@

We tested the three sampling methods on a real community time-series, from Lake Varese (IT). Lake Varese is a small (14.8 km^2^), deep (z<sub>max</sub>= 26 m), monomictic lake, in the subalpine region of north-western Italy (238 m asl). It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed (ref). Nutrient status was responsible for restructuration of the lake communities across trophic levels (e.g., Ceccuzi et al 2011, Bruel et al, 2018). Air temperature is now driving changes in plankton communities (Bruel et al, 2018).

@@

Random sampling perform the worst, as changepoint analysis was left to chance. Regular sampling provided good estimates from 8 samples (0-2 sample of absolute error). Iterative sample, where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necesserary to get the true changepoint

```{r case study read data output, include=FALSE}
myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese.txt"))
myoutput$method_f <- factor(myoutput$method, levels=c('Random', 'Regular', 'Iterative'))

@@

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(gridExtra)
p1 <- ggplot(myoutput, aes(final_n,diff_real)) + 
  geom_point(alpha = .3) +
  facet_wrap(~method_f) + theme_bw() +
  labs(title="a",x="", y="Distance to real changepoint")
p2 <- ggplot(myoutput, aes(rep(0, nrow(myoutput)),diff_real))  +
  facet_wrap(~method_f) + 
  geom_violin() +
  geom_dotplot(binaxis='y', stackdir='center', dotsize=.3) + 
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + theme_bw() +
  labs(title="b",x="", y="Distance to real changepoint")
grid.arrange(p1, p2, nrow = 2)
```

@@

>>>>

---
bibliography: Whitebib.bib
csl: ecology-letters.csl
editor_options:
  chunk_output_type: console
fontsize: 12pt
geometry: margin=1in
header-includes: \usepackage{float} \usepackage{lineno} \usepackage{setspace}\doublespacing
  \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{,} \usepackage{color} \usepackage{totcount}
  \newtotcounter{citenum} \def\oldcite{} \let\oldcite=\bibcite \def\bibcite{\stepcounter{citenum}\oldcite}
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
---

@@

Title: Sampling requirements to detect ecosystem change

\vspace{7 mm}

Authors: Rosalie Bruel and Easton R. White

\vspace{5 mm}

Addresses: \normalsize{\indent $^{1}$Department of Biology, University of Vermont, VT 05405, USA}

\vspace{5 mm}
Keywords: {time-series} \vspace{10 mm}

\vspace{5 mm}

Number of words: XXXX \vspace{5 mm}

Number of references:

Number of figures and panels: X

\vspace{5 mm}

Submission as:

\vspace{5 mm}

Preprint available at:

\clearpage

\begin{center}
\textbf{\Large Title here}
\vspace{7 mm}

Authors: \textsc{Rosalie Bruel and Easton R. White}
\vspace{5 mm}

Addresses: {\indent $^{1}$Department of Biology, University of Vermont, VT 05405, USA}

\end{center}

\textbf{Abstract}

Abstract here

\vspace{3 mm}

Keywords: {time series}

@@

- writing intro
- writing methods/results pair of section
- writing discussion
- setting up bibliography system
- Possible journals: Methods Ecology and Evolution, Biological Conservation, *Journal of Environmental Management*, *Environmental Monitoring and Assessment*, BioScience

@@

Environmental monitoring is one of the core components to modern research and management (citations). Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions (citations). Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available (Magurran2010). However, environmental monitoring is still often expensive and time-consuming (citations). Therefore, monitoring programs need to be designed in such a way to address the question of interest. Key considerations of an optimal monitoring program include the frequency of surveys, the selection of sites, as well as the precision and accuracy of the monitoring itself (citation). Monitoring programs also need to be designed to be cost-effective given limited bugets (citations).

Accounting for these considerations in the design of a monitoring program is necessary to detect long-term ecological change. The specifics of the monitoring program will determine the power with which a question of interest can be addressed. For example, @White2019 found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time. Other work has focused on the frequency of monitoring (citations), the allocation of resources spatially versus temporally (citations), and the ability to use citizen science to increase monitoring (citations). Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection (citation).

Because of new technological advances, there are many data sources that can be derived long after the actual  processes occurred. For example, paleoecological lake cores can be extracted and historical ecological communities can be reconstructured (citations from Rosalie). Similarly, environmental samples (e.g. water, soil) can be saved with eDNA processed later (citation). Similarly, video-graphical approaches can record snapshots of a system and be analyzed later (citations). In each of these cases, decisions have to made about how much data to extract from the previously collected samples. Should the paleoecological core be analyzed at every centimeter? Should the video be assessed at every minute? As long as processsing samples is expensive, these trade-offs will remain.

Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where additional samples can be added. We focus on paleoecological core samples as one example of this type of data. We examine the situation where the goal is to detect the time at which a significant change in an ecological community occurs, a changepoint. However, our approaches are widely applicable to other questions and data types. We first investigate these tools using a simulated-based approach to As a case study, we apply these tools to the detection of change points for a set of lakes. (more from Rosalie here).

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=7.5,eval=T,fig.cap='Conceptual figure for... \\label{fig:conceptual}'}
source(paste0(getwd(),"/R/Fig_conceptual.R"))

@@

We began with a theoretical exploration of the determiners of sampling required to detect a changepoint. We modeled a simple AR-1 process with a response variable that represents either population size, biodiversity, or some metric of community composition. The model includes temporal variability ($\sigma$), autocorrelation ($\phi$), the size of the shift ($\delta$), and a specific time when the shift occurred ($\tau$):

@@

We tested the three sampling methods on a real community time-series, from Lake Varese (IT). Lake Varese is a small (14.8 km^2^), deep (z<sub>max</sub>= 26m), monomictic lake, in the subalpine region of north-western Italy (238m asl). It underwent hyper-eutrophication over the 20^th^ century due to increase in nutrient loads from the watershed (ref). Nutrient status was responsible for restructuration of the lake communities across trophic levels (e.g., Ceccuzi et al 2011, Bruel et al, 2018). Air temperature is now driving changes in plankton communities (Bruel et al, 2018).

@@

Random sampling perform the worst, as changepoint analysis was left to chance. Regular sampling provided good estimates from 8 samples (0-2 sample of absolute error). Iterative sample, where 5 samples are initially sampled, then other samples are slowly added, performed the best, as no more than 9 samples were ever necessary to get the true changepoint

```{r case study read data output, include=FALSE,eval=F}
myoutput <- read.table(paste0(getwd(),"/Output/output_changepoint_varese.txt"))
myoutput$method_f <- factor(myoutput$method, levels=c('Random', 'Regular', 'Iterative'))

@@

```{r echo=FALSE, message=FALSE, warning=FALSE,eval=F}
library(ggplot2)
library(gridExtra)
p1 <- ggplot(myoutput, aes(final_n,diff_real)) + 
  geom_point(alpha = .3) +
  facet_wrap(~method_f) + theme_bw() +
  labs(title="a",x="", y="Distance to real changepoint")
p2 <- ggplot(myoutput, aes(rep(0, nrow(myoutput)),diff_real))  +
  facet_wrap(~method_f) + 
  geom_violin() +
  geom_dotplot(binaxis='y', stackdir='center', dotsize=.3) + 
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + theme_bw() +
  labs(title="b",x="", y="Distance to real changepoint")
grid.arrange(p1, p2, nrow = 2)
```

@@

\clearpage
